Headgroup Structure and Cation Binding in Phosphatidylserine Lipid Bilayers

### 2019-09-12 Manuscript/manuscriptPS.tex

\section{Results and Discussion}

\subsection{Headgroup and glycerol backbone C--H bond order parameters of POPS from $^{13}$C NMR}
The INEPT and 2D R-PDLF experiments from POPS samples gave well resolved spectra for all the
carbons in the headgroup and glycerol backbone regions (Fig. \ref{PShgSIGNSsimpson}).
The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13}, whereas
the peaks for $\beta$ and $\alpha$ carbons were assigned according to the
known C--H bond order parameters from the $^2$H\,NMR experiments~\cite{browning80}.
Slices of the R-PDLF spectra and the resulting  $S_{\rm CH}$ values
are shown in the Supporting Information (Fig. \ref{DPslices}).

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
  \caption{\label{PShgSIGNSsimpson}
    The headgroup and glycerol backbone region of the (A) Refocused-INEPT spectrum and
    (B) 2D R-PDLF spectra.
    (C) Experimental S-DROSS data (points), and SIMPSON simulations (blue lines) with
    the C--H bond order parameter values of $-0.12$ for the $\beta$-carbon, and $+0.09$ and $-0.02$
    for the $\alpha$-carbon.
    Dashed gray line is the S-DROSS curve from a SIMPSON simulation with a positive value ($+0.02$) 
    for the smaller $\alpha$-carbon C--H bond order parameter.
  }
\end{figure}
\begin{figure}[!htb]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
  \caption{\label{HGorderParameters}
    (A) Chemical structures and labels for the headgroup and glycerol backbone carbons.
    (B) Headgroup and glycerol backbone order parameters of POPS (T = 298 K) measured in this work compared
    with the previously published values from DOPS ($T=303$~K, $^2$H NMR, 0.1~M of NaCl) \cite{browning80} and 
    POPC  ($T=300$~K, $^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
    are measured in this work whereas signs of the PC order parameters are measured previously~\cite{ferreira16}.
    The size of errorbars ($\pm$0.02) shown for $^{13}$C NMR data is justified previously~\cite{botan15,ollila16}. 
  }
  %\todo{Replacing the "literature" with "from Ref X" has been suggested  https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/34.
  %We can do this just before submission when citation numbers will not change anymore.}
  %SAMULI: Maybe better to do in revision stage because citations might still change.
\end{figure}

Since the R-PDLF and previous $^2$H\,NMR experiments \cite{browning80,roux91} give 
only the absolute values of  $S_{\rm CH}$, we determined the signs of the PS headgroup
 $S_{\rm CH}$ using the S-DROSS experiment \cite{gross97}.
For a given carbon, its S-DROSS dipolar modulation profile in the indirect dimension is a superposition
of sinusoidal functions from the possible orientations of crystallites in the sample (or bilayer patches).
We phase corrected the 2D spectrum in the direct dimension such that positive and negative signs for the $S_{\rm CH}$ 
give rise to profiles that initially increase and decrease, respectively.
In practice, we use the known negative sign of the acyl chain carbons as a reference to perform
the phase correction and interpret the distinct initial slopes of the S-DROSS profiles (Fig. \ref{DPslices}). 
The S-DROSS slice for the $\beta$-carbon clearly shows an initial decrease and therefore its order parameter must be negative.
For the $\alpha$-carbon such analysis is not as trivial due to the two inequivalent order parameters of the two distinct C--H bonds.
However, the beginning of its S-DROSS slice suggests that the larger $S_{\rm CH}$ of the $\alpha$-carbon is positive and the
decrease towards negative values at longer $t_1$ suggests that the smaller  $S_{\rm CH}$ is negative.    
This is confirmed by a SIMPSON simulation
using the  $S_{\rm CH}$ values of $+0.09$ from the dipolar coupling measured here (Fig. \ref{DPslices})
and $-0.02$ from the previous $^2$H\,NMR experiment~\cite{roux91}.
We used the literature value for the smaller  $S_{\rm CH}$ , because the
resolution of our R-PDLF experiment was not sufficient to determine the
magnitude of the small value.
The S-DROSS curve from the SIMPSON simulation with a positive value for the smaller  $S_{\rm CH}$ 
(dashed grey in Fig. \ref{PShgSIGNSsimpson} C)) did not agree with the experiment, 
corroborating the interpretation that the smaller  $S_{\rm CH}$  is negative.

The headgroup and glycerol backbone order parameters of 
POPS measured in this work are in good agreement with the previously reported
values from $^2$H\,NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
When compared with the previously measured values for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}),
the $\beta$-carbon  $S_{\rm CH}$ is significantly more negative and $\alpha$-carbon
experiences a substantial forking (different  $S_{\rm CH}$ for the two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
These features have been intepreted to arise from a rigid PS headgroup
conformation, stabilized by hydrogen bonds or electrostatic
interactions \cite{browning80,buldt81}, but a detailed structrural interpretation is not
available.

We note that the DOPS $^2$H~NMR reference data found in the literature~\cite{browning80,roux90} was obtained by first solvating
the lipids with a buffer solution and then centrifuging the sample to a pellet that was used for the measurements. Such samples have a lower lipid concentration
(approximately 10~wt~\% of lipids~\cite{browning80,roux88,roux90}) than 
the gravimetric samples (60~wt~\%) and simulations (approximately 50--60~wt~\%) in this work.
Larger multilamellar repeat distances are expected in the samples with lower lipid
concentrations due to the swelling caused by electrostatic repulsion in pure PS lipid systems~\cite{millman82}.
Yet the PS headgroup  $S_{\rm CH}$  measured from gravimetric samples (POPS) in this work
are in good agreement with the results from centrifuged samples~\cite{browning80}. This, together with the rapid decrease of equilibrium repeat distance with addition of monovalent salt~\cite{millman82,rand89}, indicates that the hydration levels of multilamellae are sufficiently similar in the simulations and reference experiments.

>>>>

\section{Results and discussion}

\subsection{Headgroup and glycerol backbone C--H bond order parameters of POPS from $^{13}$C NMR}
The INEPT and 2D R-PDLF experiments from POPS samples gave well resolved spectra for all the
carbons in the headgroup and glycerol backbone regions (Fig. \ref{PShgSIGNSsimpson}).
The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13}, whereas
the peaks for $\beta$ and $\alpha$ carbons were assigned according to the
C--H bond order parameters known from previous $^2$H\,NMR experiments~\cite{browning80}.
Slices of the R-PDLF spectra and the resulting  $S_{\rm CH}$ values
are shown in the Supporting Information (Fig. \ref{DPslices}).

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
  \caption{\label{PShgSIGNSsimpson}
    The headgroup and glycerol backbone region of the (A) Refocused-INEPT spectrum and
    (B) 2D R-PDLF spectra.
    (C) Experimental S-DROSS data (points), and SIMPSON simulations with
    the C--H bond order parameter values of $-0.12$ for the $\beta$-carbon and $+0.09$ and $-0.02$
    for the $\alpha$-carbon (blue lines).
    Dashed gray line is the S-DROSS curve from a SIMPSON simulation with a positive value ($+0.02$) 
    for the smaller $\alpha$-carbon C--H bond order parameter.
  }
\end{figure}
\begin{figure}[!htb]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
  \caption{\label{HGorderParameters}
    (A) Chemical structures and labels for the headgroup and glycerol backbone carbons.
    (B) Headgroup and glycerol backbone order parameters of POPS (T = 298 K) measured in this work compared
    with the previously published values from DOPS ($T=303$~K, $^2$H NMR, 0.1~M of NaCl) \cite{browning80} and 
    POPC  ($T=300$~K, $^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
    are measured in this work, whereas signs of the PC order parameters are measured previously~\cite{ferreira16}.
    The size of errorbars ($\pm$0.02) shown for $^{13}$C NMR data is justified previously~\cite{botan15,ollila16}. 
  }
  %\todo{Replacing the "literature" with "from Ref X" has been suggested  https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/34.
  %We can do this just before submission when citation numbers will not change anymore.}
  %SAMULI: Maybe better to do in revision stage because citations might still change.
\end{figure}

Since the R-PDLF and previous $^2$H\,NMR experiments \cite{browning80,roux91} give 
only the absolute values of  $S_{\rm CH}$, we determined the signs of the PS headgroup
 $S_{\rm CH}$ using the S-DROSS experiment \cite{gross97}.
For a given carbon, its S-DROSS dipolar modulation profile in the indirect dimension is a superposition
of sinusoidal functions from the possible orientations of crystallites in the sample (or bilayer patches).
We phase corrected the 2D spectrum in the direct dimension such that positive and negative signs for the $S_{\rm CH}$ 
give rise to profiles that initially increase and decrease, respectively.
In practice, we use the known negative sign of the acyl chain carbons as a reference to perform
the phase correction and interpret the distinct initial slopes of the S-DROSS profiles (Fig. \ref{DPslices}). 
The S-DROSS slice for the $\beta$-carbon clearly shows an initial decrease and therefore its order parameter must be negative.
For the $\alpha$-carbon such analysis is not as trivial due to the two inequivalent order parameters of the two distinct C--H bonds.
However, the beginning of its S-DROSS slice suggests that the larger $S_{\rm CH}$ of the $\alpha$-carbon is positive and the
decrease towards negative values at longer $t_1$ suggests that the smaller  $S_{\rm CH}$ is negative.    
This is confirmed by a SIMPSON simulation
using the  $S_{\rm CH}$ values of $+0.09$ from the dipolar coupling measured here (Fig. \ref{DPslices})
and $-0.02$ from the previous $^2$H\,NMR experiment~\cite{roux91}.
We used the literature value for the smaller  $S_{\rm CH}$, because the
resolution of our R-PDLF experiment was not sufficient to determine the
magnitude of the small value.
The S-DROSS curve from a SIMPSON simulation with a positive value for the smaller  $S_{\rm CH}$ 
(dashed grey in Fig. \ref{PShgSIGNSsimpson} (C)) did not agree with the experiment, 
corroborating the interpretation that the smaller  $S_{\rm CH}$ is indeed negative.

The headgroup and glycerol backbone order parameters of 
POPS measured in this work are in good agreement with the previously reported
values from $^2$H\,NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
When compared with the previously measured values for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}),
the $\beta$-carbon  $S_{\rm CH}$ is significantly more negative, and the $\alpha$-carbon
experiences a substantial forking (different  $S_{\rm CH}$ for the two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
These features have been intepreted to arise from a rigid PS headgroup
conformation, stabilized by hydrogen bonds or electrostatic
interactions \cite{browning80,buldt81}, but a detailed structrural interpretation is not
available.

We note that the DOPS $^2$H~NMR reference data found in the literature~\cite{browning80,roux90} was obtained by first solvating
the lipids with a buffer solution and then centrifuging the sample to a pellet that was used for the measurements. Such samples have a lower lipid concentration
(approximately 10~wt~\% of lipids~\cite{browning80,roux88,roux90}) than 
the gravimetric samples (60~wt~\%) and simulations (approximately 50--60~wt~\%) in this work.
Larger multilamellar repeat distances are expected in the samples with lower lipid
concentrations due to the swelling caused by electrostatic repulsion in pure PS lipid systems~\cite{millman82}.
Yet the PS headgroup  $S_{\rm CH}$  measured from gravimetric samples (POPS) in this work
are in good agreement with the results from centrifuged samples~\cite{browning80}. This, together with the rapid decrease of equilibrium repeat distance with addition of monovalent salt~\cite{millman82,rand89}, indicates that the hydration levels of multilamellae are sufficiently similar in the reference experiments and our simulations.
